Clustering time series based on probability distributions across temporal granularities

### 2021-11-17 main.Rmd
start of 4.3

Clustering of the $24$ prototypes <!--and the data subset consisting of $353$ customers--> was conducted with all three distances, JS-NQT, JS-RS and WPD, and is summarised in Figures \ref{fig:opt-cluster-tsne-jsd}, \ref{fig:groups-4and5} and \ref{fig:parcoord-application}. Groups are colored differently whenever they represent different groupings. While plotting the clusters, RS-based scaling is applied to each customer, so that no customer dominates the shape of the summarized clusters because of magnitude. The plotting scales are not displayed since we want to emphasize comparable patterns rather than scales. The idea is that a customer in a cluster may have low total energy usage, but their behavior may be quite similar to a customer with high usage with respect to distributional pattern or significance across cyclic granularities.

@@

The $sindex$ is plotted against the number of clusters for clustering based on JS-based (JS-NQT) distances in Figure \ref{fig:opt-cluster-tsne-jsd}. The $sindex$ falls from around $0.56$ when $k=3$ to $0.54$ when $k$ increases. In our study, we use $k=4, 5$. Figure \ref{fig:groups-4and5} depicts the summarized distributions across $4$ and $5$ clusters in (a) and (b) respectively, and assists in characterizing each cluster. In the following plots, the median is shown by a line, and the shaded region shows the area between the  $25^{th}$ and $75^{th}$ percentiles. Groups A-1, A-2, and A-3 have profiles that correspond to B-1, B-2, and B-3 respectively. B-2 (id: 4-9 in Figure \ref{fig:hod-ind-group-png}) and B-1 (id: 21-24) have a $hod$ pattern with a typical morning and evening peak, whereas the other groups have no distinct pattern or high variability throughout the day. A-4 is subdivided into B-4 and B-5, each of which has a distinct shape across $moy$ and $wnwd$. B-4 (id: 16-20) is distinguished by higher energy consumption in the first few months of the year and greater variability in weekend usage. B-5 (id: 1-3) has higher consumption in the middle of the year (winter months) and a similar weekday-weekend inter-quartile range. When $k = 4$ is used, these two groups merge to form A-4, which has a $moy$ profile of higher usage in both the beginning and middle of the year, which is not representative of the individuals. It may be worthwhile to compare Figures \ref{fig:opt-cluster-tsne-jsd} and \ref{fig:hod-ind-group-png} to see if the summarized distributions across groups accurately characterized the groupings. If it has, then the majority of the members of the group should have a similar profile.

@@

```{r opt-cluster-tsne-jsd, fig.cap = "Clustering summaries: t-SNE computed on the $24$ selected customers (a), and separation index ($sindex$) for $2$-$10$ clusters using JS-NQT, JS-RS and WPD in (b). Various choices in number of clusters would be recommended. Four clusters are visible in t-SNE, although it might hide a fifth cluster because dimension reduction to 2D may be insufficient to see the difference. JS-NQT suggests $3$, JS-RS suggests $3$ or $5$ and WPD suggests $3$ or $6$."}
#include_graphics("figs/opt_clust_parenthesis.png")
```

>>>>

Clustering of the $24$ prototypes <!--and the data subset consisting of $353$ customers--> was conducted with all three distances, JS-NQT, JS-RS and WPD, and is summarised in Figures \ref{fig:opt-cluster-tsne-jsd}, \ref{fig:groups-4and5} and \ref{fig:parcoord-application}. The t-SNE visualization suggests that there are four well-separated clusters. It is possible that because the representation is only 2D, that the fifth group from the original prototype selection is distinctly different in high dimensions. The $sindex$ plots for the three methods indicate some disagreement: JS-NQT suggests 3, JS-RS suggests 3 or 5 and WPD suggests 3 or 6.  JS-RS would appear to match the original prototypes with the five cluster solution, but it actually differs. Even though the $sindex$ for JS-NQT suggests three clusters, the five cluster solution more closely matches the original prototypes. The WPD clustering provides a different grouping of the customers, and even though it disagrees with the original prototypes it is a useful grouping.  <!--While plotting the clusters, RS-based scaling is applied to each customer, so that no customer dominates the shape of the summarized clusters because of magnitude. The plotting scales are not displayed since we want to emphasize comparable patterns rather than scales. The idea is that a customer in a cluster may have low total energy usage, but their behavior may be quite similar to a customer with high usage with respect to distributional pattern or significance across cyclic granularities.-->

@@

Figure \ref{fig:opt-cluster-tsne-jsd} The $sindex$ is plotted against the number of clusters for clustering based on JS-based (JS-NQT) distances in Figure \ref{fig:opt-cluster-tsne-jsd}. The $sindex$ falls from around $0.56$ when $k=3$ to $0.54$ when $k$ increases. In our study, we use $k=4, 5$. Figure \ref{fig:groups-4and5} depicts the summarized distributions across $4$ and $5$ clusters in (a) and (b) respectively, and assists in characterizing each cluster. In the following plots, the median is shown by a line, and the shaded region shows the area between the  $25^{th}$ and $75^{th}$ percentiles. Groups A-1, A-2, and A-3 have profiles that correspond to B-1, B-2, and B-3 respectively. B-2 (id: 4-9 in Figure \ref{fig:hod-ind-group-png}) and B-1 (id: 21-24) have a $hod$ pattern with a typical morning and evening peak, whereas the other groups have no distinct pattern or high variability throughout the day. A-4 is subdivided into B-4 and B-5, each of which has a distinct shape across $moy$ and $wnwd$. B-4 (id: 16-20) is distinguished by higher energy consumption in the first few months of the year and greater variability in weekend usage. B-5 (id: 1-3) has higher consumption in the middle of the year (winter months) and a similar weekday-weekend inter-quartile range. When $k = 4$ is used, these two groups merge to form A-4, which has a $moy$ profile of higher usage in both the beginning and middle of the year, which is not representative of the individuals. It may be worthwhile to compare Figures \ref{fig:opt-cluster-tsne-jsd} and \ref{fig:hod-ind-group-png} to see if the summarized distributions across groups accurately characterized the groupings. If it has, then the majority of the members of the group should have a similar profile.

@@

```{r opt-cluster-tsne-jsd, fig.cap = "Clustering summaries:  (a) t-SNE computed on the $24$ selected customers, and  (b) separation index ($sindex$) for $2$-$10$ clusters using JS-NQT, JS-RS and WPD. Various choices in number of clusters would be recommended. Four clusters are visible in t-SNE, although it might hide a fifth cluster because dimension reduction to 2D may be insufficient to see the difference. JS-NQT suggests $3$, JS-RS suggests $3$ or $5$ and WPD suggests $3$ or $6$."}
#include_graphics("figs/opt_clust_parenthesis.png")
```
